---
title: "Homework 7 - Ensemble Random Forest"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Dataset and Objective

The dataset consists of [Phishing Websites](https://www.kaggle.com/akashkr/phishing-website-dataset) from Kaggle. It uses 30 different features to determine whether or not a given website is a phishing website. A phishing website is a website which tries to steal information from users under the guise of looking and behaving like a legitimate website.

```{r message=FALSE, warning=FALSE}
library(caret)
set.seed(408)

phishing <- read.csv('phishing-dataset.csv')

# Row id not needed
phishing <- phishing[,-1]

# Factorize remaning columns
phishing[,] <- lapply(phishing[,], factor)
summary(phishing$Result)
```

There are 4,898 instances of phishing websites and 6,157 legitimate websites My objective is to build a classifier that can use the 30 features given to determine the website class as phishing or legitimate.

### Explanation of features

```{r message=FALSE, warning=FALSE}

data.frame(t(phishing[1,]))

```

The model has 30 features, below is a description of some and how they relate to whether or not the URL should be classified as a phishing site. For each feature, a value of 1 indicates legitimate, 0 indicates suspicious and -1 indicates phishing.

__URL_Length__ - Between 54 and 75 is consideredsuspicious. Greater than 75 is classified as phishing.

__Shortening_Service__ - Was a shortening service used to conceal the true length of the URL?

__SSLfinal_State___ - If https is used, is the certificate issuer trusted? If not it's suspicious. If https is not used, it's considered phishing.

__Favicon__ - Was the favicon loaded from a different domain?

__URL_of_Anchor__ - Do more than 67% of of the `<a>` not contain any links? If so, it's classified as phishing. If greater than 37% but less than 67% of the `<a>` tags do not contain links, then it's considered suspicious.

__SFH__ - Is form data submitted to an external domain or is no action taken upon submission?

__Submitting_to_email__ - Are form submissions emailed?

__on_mouseover__ - Is JavaScript used to change the status bar in the browser?

__web_traffic__ - Does the website have a rank greater than 100,000 in the Alex database? The latter database ranks the popularity of a website by the number of visitors in addition to the number of websites they visit.

__Google_Index__ - Has the site been indexed by Google?

__Statistical_report__ Does the IP belong to a known list of phishing IPs?

### Training the Random Forest

```{r echo = FALSE}
rfModel <- readRDS("./rfModel.rds")

trainIndex <- createDataPartition(phishing$Result, p = 0.75, list=FALSE)
trainData <- phishing[trainIndex,]
trainTest <- phishing[-trainIndex,]
trControl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
```

```{r message=FALSE, warning=FALSE, eval=FALSE}
trainIndex <- createDataPartition(phishing$Result, p = 0.75, list=FALSE)
trainData <- phishing[trainIndex,]
trainTest <- phishing[-trainIndex,]
trControl <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

rfModel <- train(Result ~ ., data = trainData, method = "rf", metric = "Kappa", trControl = trControl, tuneGrid = data.frame(mtry = c(1, 2, 4, 8, 16)))
```

```{r message=FALSE, warning=FALSE}
rfModel

confusionMatrix(rfModel)

plot(rfModel)
```

The model with the highest accuracy and kappa is one which selected from 16 candidate features at each split.

### Classifying the test data

```{r}
predictions <- predict(rfModel, newdata=trainTest[,-ncol(trainTest)])

confusionMatrix(predictions, trainTest$Result)

```

```{r}
library(ggplot2)

rfVarImp <- varImp(rfModel, scale = TRUE)

rfVarImp <- data.frame(rownames(rfVarImp$importance), rfVarImp$importance$Overall)

names(rfVarImp) <- c('Website', 'Importance')

rfVarImp <- transform(rfVarImp, Website = reorder(Website, Importance))

ggplot(data=rfVarImp, aes(x=Website, y=Importance)) + 
  geom_bar(stat = 'identity',colour = "blue", fill = "blue") + 
  coord_flip()

```

The features which are the highest predictors in classifying a phishing website, according to this model, are whether or not the website uses SSL/TLS and has a trusted certificate, the percentage of anchor tags (`<a>`) on the page which contain links and the number of visitors to the website.